Komols Complexity, Coresets, and Sketches in Machine Learning

### 2018-12-28 komlos.tex
commit before pull

\newcommand{\ip}[1]{\left \langle #1 \right \rangle}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\D}{\rho}
\newcommand{\eps}{\epsilon}
\newcommand{\F}{\mathcal{F}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Q}{\mathcal{Q}}
%\usepackage{ntheorem}
\newcommand{\floor}[1]{\left \lfloor #1 \right \rfloor}
\newcommand{\ceil}[1]{\left \lceil #1 \right \rceil}
\newcommand{\disc}{\text{disc}}

@@

\begin{abstract}
This paper presents the Komlos complexity for families of functions.
We argue that this measure of complexity captures properties like minimal coreset sizes, streaming mergable sketches, and batch active learning.  
We show that for many well known problems the Komlos complexity is asymptotically lower than their Rademacher complexity.
This proves the existence smaller coresets than those achievable by random sampling. 
These include frequent item counting, quantile approximation, kernel density estimation, matrix covariance approximation, and linear and logistic regression.
For all of the above we match the state of the art or improve on it with the above generic framework.
This makes an explicit connection between low discrepancy, streaming algorithms, coresets, and learnability. 
We also asymptotically improve on the generic merge-reduce framework for creating mergeable sketches for all the above problems. 
\end{abstract}

@@

Considering a random i.i.d.\ Gaussian Tensor $g_k \in \R^{d^k}$. 
A trivial modification of Theorem 1 from \cite{tomioka2014spectral} shows that $\Pr[\|g_k\| \ge c\sqrt{d\log(k)}] \le 1/10k^2$ for some constant $c$. The only change needed in the proof is the size of the epsilon net which changes from $(2\log(3/2)/k)^{kd}$ for \cite{tomioka2014spectral} to $(2\log(3/2)/k)^d$. Union bounding on all values of $k$ we get $\sum_k 1/10k^2 \le 1/2$ which shows $g = [g_1, \operatorname{flat}(g_2), \operatorname{flat}(g_3), \ldots]$ belongs to $\mathcal K$ with probability al least $1/2$. The function $\operatorname{flat}(X)$ simply strings the entry values of the tensor $X$ into a vector.

We now define a mapping $\psi(x)$ of $x\in \R^d$ to a high dimensional space. 
$$\psi(x) = [x, \frac{\operatorname{flat}(x^{\otimes 2})}{\sqrt{2\log^2(2)}}, \frac{\operatorname{flat}(x^{\otimes 3})}{\sqrt{3\log^2(3)}}, \ldots,\frac{\operatorname{flat}(x^{\otimes k})}{\sqrt{k\log^2(k)}},\ldots]$$
Finally, note that for $\|x\| \le 1$ we have $\|\psi(x)\| = (\sum_k  1/k\log^2(k))^{1/2} = O(1)$.

We are now ready to apply Banaszczyk's theorem. 
There exist signs $\sigma_i$ such that $\psi  = \sum_i \sigma_i \psi(x_i) \in C \mathcal K$.
Since $\psi_k = \sum_i \sigma_i x_i^{\otimes k}/\sqrt{k \log^2{k}}$ we get that $\max_k \|\sum_i \sigma_i  x_i^{\otimes k}\| \le \sqrt{d k \log^{3}(k)}$

@@

In \cite{DBLP:journals/corr/abs-1802-01751} (see paragraph before Lemma 2.2) they manage to obtain a coreset (with equal weights) of size $\sqrt{d \log(1/\eps)}/\eps$ for any kernel with certain properties. They do so by considering a matrix whose rows correspond to all the vectors in an $\eps$-net over the space, and columns correspond to the $n$ data points. Calling that matrix $A$, $A_{q,x} = K(q,x)$. They bound the discrepancy of the matrix defined as $\min_s \|As\|_\infty$ where $s$ is a sign vector by using a bound 
$$ \min_s \|As\|_\infty \leq \sqrt{\log(N)} \|A\|_{\gamma_2} $$
Here, $N$ is the size of the $\eps$ net and the $\gamma_2$ norm is 
$$ \|A\|_{\gamma_2} = \min_{B,C: A=BC^T} \text{max norm of a row of B or C}$$
They use the fact that for psd kernel $K$, $\|A\|_{\gamma_2} \leq 1$ (in fact it is equal to 1). Now, back to our case. For any fixed power $k$ we know that $K(x,q)=\ip{x,q}^k$ is a psd kernel, hence its corresponding matrix $A_k$ has $\gamma_2$ norm bounded by $1$. Now, the logistic function can be written in a taylor expansion whose coefficients, in absolute value, sum to a constant $C$. It follows that the $A$ matrix for the logistic function has a $\gamma_2$ norm bounded by $C$. The same arguments as in  \cite{DBLP:journals/corr/abs-1802-01751} can now prove that its discrepancy is bounded by $\sqrt{\log(N)} \approx \sqrt{d\log(1/\eps)}$

A useful fact about the $\gamma_2$ norm is that it is indeed a norm. This is proven in \cite{tomczak1989banach}. In \cite{matouvsek2014factorization} this is listed in Section 2, Proposition 2.2 along with other known properties of the $\gamma_2$ norm.

@@

%
%\subsection{notes about row selection}
%The problem is 
%$$ \min_s \| \sum_i \sigma_i x_ix_i^T\| $$
%where $\|x_i\| \leq 1$. Consider the following greedy approach: At point $x_i$, let
%$$ C_i = \sum_{j<i} \sigma_j x_j x_j^T, \ \  l_+ = \|C_i + x_i x_i^T \|, \ l_- = \|C_i - x_i x_i^T\|$$
%We choose $\sigma_i = 1$ if $l_+ < l_i$, otherwise $\sigma_i=-1$.
%
%This approach can be shown to be better than random. Random will result in an answer of up to 
%
%Consider the following example. 
%$$x_1=e_1, x_2=e_2*\eps, x_3=e_2, x_4=e_1/\sqrt{2} + e_2/\sqrt{2}$$
%where $\eps \to 0$. We get that the covariance matrix after inserting $x_3$ has two eigenvalues of +1 and -1. Now when inserting $x_4$ we have to increase the magnitude of one of them and we exceed the limit of 1. 
%

\section{Coreset for linear functions (based on $\gamma_2$ norm)}

Let $x_1,\ldots,x_n \in R^d$ be and $y_1,\ldots,y_n$ be labels associated to the data points. Let $K$ be a psd kernel. Consider a problem of learning a linear function aiming to minimize the loss of 
$$\min_q \sum_{i=1}^n L(K(x_i, q), y_i)$$

Our objective is to find a core-set of the $n$ data points. That is, we aim to find a set $S \subseteq [n]$ such that for every $q$ it holds that 
$$ \left| \sum_{i=1}^n L(K(x_i, q), y_i) - \frac{n}{|S|} \sum_{i \in S} L(K(x_i, q), y_i) \right| \leq \eps n $$

For a function $f(q,x)$, let $A_f$ be a matrix whose rows are an $\eps$-net over the Euclidean unit ball of $R^d$, and columns correspond to the different $x_i$'s. The value of $A_f$ at an entry corresponding to $(q,x)$ if $f(q,x)$.

\begin{definition}
For a matrix $A$,
$$ \|A\|_{\gamma_2} = \min_{B,C: A=BC^T} \|B\|_{2 \to \infty} \|C\|_{2 \to \infty}$$
is the $\gamma_2$ norm of $A$. 
\end{definition}
 \cite{tomczak1989banach} prove that this is indeed a norm, in particular that 
 $\|A_1+A_2\|_{\gamma_2} \leq \|A_1\|_{\gamma_2} + \|A_2\|_{\gamma_2}$.

\begin{lemma} \label{lem:powers}
For $k \geq 0$ let $f_k(q,x) = \ip{q,x}^k$. If holds that $\|A_{f_k}\|_{\gamma_2} \leq 1$
\end{lemma}
\begin{proof}
For $k=0$ the statement is obvious. For $k>0$, notice that $f_k$ is a kernel, and in particular there exists a mapping $\phi_k: \R^d \to \R^{d^k}$ such that
$$ \ip{q,x}^k = \ip{\phi_k(q), \phi_k(x)} $$
Furthermore, 
$$ \|\phi_k(x)\| = \sqrt{\ip{\phi_k(x), \phi_k(x)}} = \sqrt{\|x\|^{2k}} = \|x\|^k \leq 1 $$
Since all $x,q$ have at most unit norm we get a bound over the $\gamma_2$ norm by viewing the $B,C$ matrices obtained from the $\phi_k$ embeddings of the different $x,q$'s
\end{proof}

\begin{lemma}
Let $L:\R \to \R$ be a function that can be written as 
$$ L(z) = \sum_{k=0}^\infty \alpha_k z^k$$
with $\sum_{k=0}^{\infty} |\alpha_k| \leq C$. For $f_L(q,x) = L(\ip{q,x})$ it holds that 
$$ \|A_{f_L}\|_{\gamma_2} \leq C $$
\end{lemma}
\begin{proof}
This is a direct corollary from $\gamma_2$ being a norm, and Lemma~\ref{lem:powers}
\end{proof}

This lemma in particular holds for the logistic function as its Taylor series has elements that converge. In particular, evaluating the function at $i$ and summing the real and imaginary parts of the result gives the sum of the absolute values of the coefficients, which is (roughly) $0.461$
\begin{corollary}
For $L(z) = \log(1+e^z)$ we have $\|A_{f_L}\|_{\gamma_2} \leq 0.5$
\end{corollary}

\begin{definition}
For a matrix $A$ of dimensions $m \times n$, $\disc(A)$ is defined as 
$$ \min_{s \in \{-1,1\}^n} \|As\|_\infty $$
\end{definition}
In \cite{matouvsek2014factorization} they prove that 
$$\|A\|_{\gamma_2} / \log(m) \leq \disc(A) \leq \|A\|_{\gamma_2}  \sqrt{\log(m)}$$
This gives a way to control the discrepancy of a matrix by representing it as a sum of matrices with known discrepency or $\gamma_2$ norm. In particular, for the logistic function we get that 
$$\disc(A_{f_L}) \leq \sqrt{\log(|Q|)} $$
where $Q$ is an $\eps$-net over the Euclidean unit sphere. This gives the bound of 
$$\disc(A_{f_L}) \leq O(\sqrt{d\log(1/\eps)}) $$

Let's discuss what this means. For a logistic regression problem, by observing the weights and labels we can change the signs of the inputs so that the logistic loss is always $\log(1+\exp(\ip{q,x}))$, and we can now convert the discrepancy result into a coreset of size $\sqrt{d\log(1/\eps)}/\eps$, as opposed to a random sample of $1/\eps^2$.

Let $K$ be a psd kernel, meaning for $x,q \in \R^d$ we have for some mapping $\phi$ into a Hilbert space $K(q,x) = \ip{\phi_K(q), \phi_K(x)}$. Notice that for $\phi_k$ defined above
$$ \ip{\phi_k(\phi_K(q)), \phi_k(\phi_K(x))} = \ip{\phi_K(q), \phi_K(x)}^k = K(q,x)^k $$
meaning that $K^k(q,x)$ is a psd kernel for all $k \geq 0$. By using the exact same arguments before we can define $A_{K,k} \in \R^{|Q|,n}$ with $A_{K,k}(q,i) = K(q, x_i)^k$ and conclude that $\|A_{K,k}\|_{\gamma_2} \leq 1$. Then, for the matrix $A_{K,f_L}$ with
$$ L(z) = \sum_{k=0}^\infty \alpha_k z^k, A_{K,f_L}(q,i) = L(K(q,x_i))$$
we have 
$$\|A_{K,f_L}\|_{\gamma_2} \leq \sum_{k=0}^{\infty} |\alpha_k| $$

\begin{corollary}
For $L(z)=\log(1+\exp(z))$ and $L(z) = z^2$, and any psd kernel $K:\R^d \times \R^d \to \R$ we have
$$\|A_{K,f_L}\|_{\gamma_2} \leq 1 $$ 
and
$$\disc(A_{K,f_L}) \leq O(\sqrt{d\log(1/\eps)}) $$ 
\end{corollary}

\bibliographystyle{plain}
\bibliography{density}

>>>>

\newcommand{\ip}[1]{\left \langle #1 \right \rangle}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\D}{\rho}
\newcommand{\eps}{\epsilon}
\newcommand{\F}{\mathcal{F}}
\newcommand{\X}{\mathcal{X}}
\newcommand{\Q}{\mathcal{Q}}
%\usepackage{ntheorem}
\newcommand{\floor}[1]{\left \lfloor #1 \right \rfloor}
\newcommand{\ceil}[1]{\left \lceil #1 \right \rceil}
%\newcommand{\Pr}{\operatorname{Pr}}

@@

\begin{abstract}
This paper defines the Komlos complexity for families of functions which is intimately related to its discrepancy properties.
We argue that this measure of complexity captures properties like minimal coreset sizes for different problems in machine learning. We show that for many well known problems the Komlos complexity is asymptotically lower than their Rademacher complexity. This proves the existence smaller coresets than those achievable by random sampling. 
New results for coresets for kernel density estimation, matrix covariance approximation, and logistic regression are presented both as examples of for providing mechanisms to compute the Komlos complexity of other classes.
For all of the above we match the state of the art or improve on it.
This makes an explicit connection between low discrepancy, streaming algorithms, coresets, and learnability. 
%We also asymptotically improve on the generic merge-reduce framework for creating mergeable sketches for all the above problems. 
\end{abstract}

@@

Considering a random i.i.d.\ Gaussian Tensor $g_k \in \R^{d^k}$. 
A trivial modification of Theorem 1 from \cite{tomioka2014spectral} shows that $\Pr[\|g_k\| \ge c\sqrt{d\log(k)}] \le 1/10k^2$ for some constant $c$. The only change needed in the proof is the size of the epsilon net which changes from $(2\log(3/2)/k)^{kd}$ for \cite{tomioka2014spectral} to $(2\log(3/2)/k)^d$. Union bounding on all values of $k$ we get $\sum_k 1/10k^2 \le 1/2$ which shows $g = [g_1, \operatorname{flat}(g_2), \operatorname{flat}(g_3), \ldots]$ belongs to $\mathcal K$ with probability al least $1/2$. The function $\operatorname{flat}(X)$ simply strings the entry values of the tensor $X$ into a vector.
%
We now define a mapping $\psi(x)$ of $x\in \R^d$ to a high dimensional space. 
$$\psi(x) = [x, \frac{\operatorname{flat}(x^{\otimes 2})}{\sqrt{2\log^2(2)}}, \frac{\operatorname{flat}(x^{\otimes 3})}{\sqrt{3\log^2(3)}}, \ldots,\frac{\operatorname{flat}(x^{\otimes k})}{\sqrt{k\log^2(k)}},\ldots]$$
Note that for $\|x\| \le 1$ we have $\|\psi(x)\| = (\sum_k  1/k\log^2(k))^{1/2} = O(1)$.
%
We are now ready to apply Banaszczyk's theorem. 
There exist signs $\sigma_i$ such that $\psi  = \sum_i \sigma_i \psi(x_i) \in C \mathcal K$.
Since $\psi_k = \sum_i \sigma_i x_i^{\otimes k}/\sqrt{k \log^2{k}}$ we get that $\max_k \|\sum_i \sigma_i  x_i^{\otimes k}\| \le \sqrt{d k \log^{3}(k)}$

@@

In \cite{DBLP:journals/corr/abs-1802-01751} (see paragraph before Lemma 2.2) they manage to obtain a coreset (with equal weights) of size $\sqrt{d \log(1/\eps)}/\eps$ for any kernel with certain properties. They do so by considering a matrix whose rows correspond to all the vectors in an $\eps$-net over the space, and columns correspond to the $n$ data points. Calling that matrix $A$, $A_{q,x} = K(q,x)$. They bound the discrepancy of the matrix defined as $\min_s \|As\|_\infty$ where $s$ is a sign vector by using a bound 
$$ \min_s \|As\|_\infty \leq \sqrt{\log(N)} \|A\|_{\gamma_2} $$
Here, $N$ is the size of the $\eps$ net and the $\gamma_2$ norm is 
$$ \|A\|_{\gamma_2} = \min_{B,C: A=BC^T} \text{max norm of a row of B or C}$$
They prove that for psd kernel $K$, $\|A\|_{\gamma_2} \leq 1$ (in fact it is equal to 1). Now, back to our case. For any fixed power $k$ we know that $K(x,q)=\ip{x,q}^k$ is a psd kernel, hence its corresponding matrix $A_k$ has $\gamma_2$ norm bounded by $1$. Now, the logistic function can be written in a taylor expansion whose coefficients, in absolute value, sum to a constant $C$. It follows that the $A$ matrix for the logistic function has a $\gamma_2$ norm bounded by $C$. The same arguments as in  \cite{DBLP:journals/corr/abs-1802-01751} can now prove that its discrepancy is bounded by $\sqrt{\log(N)} \approx \sqrt{d\log(1/\eps)}$

@@

%
%\subsection{notes about row selection}
%The problem is 
%$$ \min_s \| \sum_i \sigma_i x_ix_i^T\| $$
%where $\|x_i\| \leq 1$. Consider the following greedy approach: At point $x_i$, let
%$$ C_i = \sum_{j<i} \sigma_j x_j x_j^T, \ \  l_+ = \|C_i + x_i x_i^T \|, \ l_- = \|C_i - x_i x_i^T\|$$
%We choose $\sigma_i = 1$ if $l_+ < l_i$, otherwise $\sigma_i=-1$.
%
%This approach can be shown to be better than random. Random will result in an answer of up to 
%
%Consider the following example. 
%$$x_1=e_1, x_2=e_2*\eps, x_3=e_2, x_4=e_1/\sqrt{2} + e_2/\sqrt{2}$$
%where $\eps \to 0$. We get that the covariance matrix after inserting $x_3$ has two eigenvalues of +1 and -1. Now when inserting $x_4$ we have to increase the magnitude of one of them and we exceed the limit of 1. 
%
\bibliographystyle{plain}
\bibliography{density}
